Implementation and comparison of register allocation to translate \lstinline!LLVM--! to x86

### 2023-11-22 article.tex
More work on intro etc.

\usepackage{tikz}
\usetikzlibrary{chains}
\usepackage{graphicx} % Required for inserting images
\usepackage[T1]{fontenc}
\usepackage{beramono}
\usepackage{courier}
\usepackage{listings}
\usepackage{xpatch}
\usepackage{xcolor}
\usepackage{realboxes}
\usepackage{titlesec}

@@

Compilers can be divided into several distinct phases, with the two halfs of which colloquially referred to as the \textit{frontend} and \textit{backend} stages. With the former serving the purpose of translating a high-level programming language to an \textit{intermediate representation} (IR) layer and the latter translating said IR to executable machine code on a target platform.

\begin{figure}[h]
  \centering
\includegraphics[width=0.8\textwidth]{phases.png}
\caption{Compiler phases}
\end{figure}

@@

\section{Review of literature}

Write about Appel book.

Write about SSA form liveness analysis.

Keywords: program points, dataflow equations, fixed point

@@

The instructions included are
\begin{enumerate}
    \item binary operations \lstinline!add!, \lstinline!and!, \lstinline!ashr!, \lstinline!lshr!, \lstinline!mul!, \lstinline!or!, \lstinline!sdiv!, \lstinline!srem!, \lstinline!shl!, \lstinline!sub! and \lstinline!xor!
    \item integer comparison \lstinline!icmp! (conditions being \lstinline!eq!, \lstinline!ne!, \lstinline!sge!, \lstinline!sgt!, \lstinline!sle! and \lstinline!slt!)
    \item memory/address operations \lstinline!alloca!, \lstinline!gep!, \lstinline!load! and \lstinline!store!
    \item  \lstinline!mov! operations \lstinline!gep!, \lstinline!phi!, \lstinline!ptrtoint! and \lstinline!zext!
    \item control flow operations \lstinline!br!, \lstinline!call! and \lstinline!ret!
    \item a  nullary terminator \lstinline!unreachable!
\end{enumerate}
%with only four distinct block terminators 
%\begin{enumerate}
%    \item Unconditional \lstinline!br! followed by the label to branch to
%    \item Conditional \lstinline!br!  followed by an \lstinline!i1! condition and two labels
%    \item \lstinline!ret! followed by an optional return value
%    \item \lstinline!unreachable!
%\end{enumerate}
The branching for most of these is trivial: after successful execution the flow of all but \lstinline!br!, \lstinline!ret! and \lstinline!unreachable! will unconditionally attempt to execute the next instruction in memory (i.e. increment the instruction pointer by instruction length).

@@

For a given program, it's representation as linear sequence of instructions from the above set isn't immediately useful for liveness analysis. (FIXME: why not?)  hence a \textit{control flow graph} is derived from these. A terminator will branch elsewhere to continue execution. The destination of this has to be an annotated by a label, hence labels \textit{initiate} blocks.

@@

@@

%% TODO
%* optimistic coloring

@@

\section{Codegen}

@@

\section{Optimization}

@@

Benchmarks are performed on Apple Silicon (M1 Pro, i.e. ARMv8.5-A \cite{AM1}) despite the target instruction set being x86-64. This is possible because of Rosetta 2, which  translates x86-64 to ARMv8 before execution \cite{A2012}.  Time is measured using the \lstinline!time! function of the bash shell as it is native to the chosen Docker image.  As all execution relevant to the performance of register allocation is done in user mode, only the \lstinline!user! metric is noted.

@@

%TODO:
%
%* should I use shell time, or /usr/bin/time?
%* how do I measure cpu cycles?

@@

>>>>

\usepackage{amsmath}
\usepackage{tikz}
\usetikzlibrary{chains}
\usetikzlibrary{angles, quotes}
\usetikzlibrary{decorations.pathreplacing}
\usepackage{graphicx} % Required for inserting images
\usepackage[T1]{fontenc}
\usepackage{beramono}
\usepackage{courier}
\usepackage{listings}
\usepackage{xpatch}
\usepackage{xcolor}
\usepackage{realboxes}
\usepackage{titlesec}

@@

Compilation refers to the process of translating from one language to another, most often from a programming language intended for humans to work with, to machine-/bytecode intended to be executed on a target platform. This process can be divided into several distinct phases, which are grouped into one of two stages, colloquially referred to as the \textit{frontend} and \textit{backend}. The former translating a high-level programming language to an \textit{intermediate representation} (IR) layer and the latter translating IR to executable machine code on a target platform (e.g. x86, ARM) or bytecode for a target VM (e.g. JVM, CLR).

\begin{figure}[h]
  \centering
  \resizebox{0.8\textwidth}{!}{
\begin{tikzpicture}
  \tikzset{
box/.style={draw,
  inner sep=0.3cm,
    minimum height=1.75cm,
    align=center},
}

  % Nodes
  \node [box, left, align=center] (incg) at (0,2.25) {Intermediate\\Codegen};
  \node [box, left=0.7 of incg, align=center] (semant) {Semantic\\Analysis};
  \node [box, left=0.7 of semant, align=center] (parse) {Parser};
  \node [box, left=0.7 of parse, align=center] (lex) {Lexical\\Analysis};
  \node [ left=0.9 of lex, align=center] (entry) {};
  \node [box, left, align=center] (cfa) at (0,0) {Control-\\Flow\\Analysis};
  \node [box, left=0.7 of cfa, align=center] (lva)  {Liveness/\\Dataflow\\Analysis};
  \node [box, left=0.7 of lva, align=center] (alloc)  {Register\\Allocation};
  \node [box, left=0.7 of alloc, align=center] (x86)  {Instruction\\Selection};
  \node [ left=0.70 of x86, align=center] (exit) {};

  % Edge with right-angle corners
  \draw [-latex] (entry.east) -- (lex.west)  node [near start,pos=0.325,rotate=90,fill=white,opacity=0.9] {\small Source};
  \draw [-latex] (lex.east) -- (parse.west)  node [midway,pos=0.4,rotate=90,fill=white,opacity=0.9] {\small Tokens};
  \draw [-latex] (parse.east) -- (semant.west)  node [midway,pos=0.4,rotate=90,fill=white,opacity=0.9] {\small AST};
  \draw [-latex] (semant.east) -- (incg.west)  node [midway,pos=0.4,rotate=90,fill=white,opacity=0.9] {\small \(t\)-AST};
  \draw [-latex] (incg.east) -- +(0.4,0) |- node[pos=0.25,fill=white,opacity=0.9] {IR} (cfa.east);
  \draw [-latex] (cfa.west) -- (lva.east)  node [midway,pos=0.4,rotate=-90,fill=white,opacity=0.9] {\small Flow Graph};
  \draw [-latex] (lva.west) -- (alloc.east)  node [midway,pos=0.4,rotate=-90,fill=white,opacity=0.9] {\small Int. Graph};
  \draw [-latex] (alloc.west) -- (x86.east)  node [midway,pos=0.4,rotate=-90,fill=white,opacity=0.9] {\small Reg. Asn.};
  \draw [-latex] (x86.west) -- (exit.east)  node [midway,pos=0.4,rotate=-90,fill=white,opacity=0.9] {\small x86};

  % backend outline
  \fill [opacity=0.1] ([shift={(-0.15,-0.15)}]x86.south west) rectangle ([shift={(0.8,0.15)}]cfa.north east);

  \draw [decorate,decoration={brace,amplitude=5pt,mirror,raise=4ex}]
    (0.3,-1) -- (0.3,1) node[midway,xshift=3em,rotate=-90]{Backend};
  \draw [decorate,decoration={brace,amplitude=5pt,mirror,raise=4ex}]
    (0.3,1.25) -- (0.3,3.25) node[midway,xshift=3em,rotate=-90]{Frontend};
\end{tikzpicture}}
\caption{Compiler phases, backend highlighted}
\end{figure}
A critical part of the backend stage is assigning each variable (also referred to as \textit{virtual}) of the source program to a physical register on the target processor.

@@

\section{Review of Literature}

The two primary sources are the Tiger- \cite{AppelAndrewW} and Dragonbook \cite{alma99122774374105763}, in addition to scientific publications concerning the linear scan approach.

\subsection{Control Flow Analysis}

The backend of a compiler takes some form of IR as input, usually a linear sequence of instructions for each separate function. This representation is close to the level of an actual processor by design, but it isn't immediately useful for the further analysis steps needed to generate optimized code for the target architecture. %%(FIXME: why not?)
The control flow of a given program refers to the order in which instructions are executed. While the flow of most instructions is linear, in the sense that the next instruction to be executed is located immediately after, some send the execution elsewhere or outright terminate it. %% FIXME: why is this important?

These instructions are referred to as \textit{terminators}, as they potentially terminate a contiguous flow of execution. This subset of instructions can be further divided into \textit{branch instructions} and those concerned with handling the callstack in function/subroutine management.

These so-called \textit{branch instructions} that alter the flow of execution are further c

hence a \textit{control flow graph} is derived from these. A terminator will branch elsewhere to continue execution. The destination of this has to be an annotated by a label, hence labels \textit{initiate} blocks.

\subsection{Liveness Analysis}

Liveness analysis is the process of finding the program points at which a variable is live. This serves the function of determining which variables are live simultaneously, i.e. in interference/conflict with one another, so cannot be assigned the same physical register without overwriting the value of one another in execution. These can be derived recursively by iteratively applying so-called \textit{dataflow equations} to each node in a \textit{control-flow graph} (CFG) until a stable state/fixed point is reached.

The algorithm described in \cite{AppelAndrewW} and \cite{alma99122774374105763} is based on the following equations for the live-in and live-out variables respectively:
\begin{align}
  in\left[n\right] &= use\left[n\right] \cup (out\left[n\right] - def\left[n\right])\\
  out\left[n\right] &= \bigcup_{s\in succ\left[n\right]} in\left[s\right]
\end{align}
where \(use\left[n\right]\) is the set of all dependent variables, \(def\left[n\right]\) is the set of all variables defines and \(succ\left[n\right]\) is the set of all immediate successor nodes of \(n\).

The structure of the CFG is inherently vague as both  \cite{AppelAndrewW} and \cite{alma99122774374105763}, while based on the same unde

rlying conceptions of dataflow, have different approaches to nodes of a CFG, with the former taking each individual instruction into consideration and the latter each basic block. Either of these approaches are sound

% TODO: try int set instead of symbols

\subsection{Graph Coloring}

\subsection{Linear Scan}

%Write about Appel book.

%Write about SSA form liveness analysis.

%Keywords: program points, dataflow equations, fixed point

@@

The instructions included are
\lstinline!trunc! has only been added in order to help cover more generated LLVM.
%with only four distinct block terminators 
%\begin{enumerate}
%    \item Unconditional \lstinline!br! followed by the label to branch to
%    \item Conditional \lstinline!br!  followed by an \lstinline!i1! condition and two labels
%    \item \lstinline!ret! followed by an optional return value
%    \item \lstinline!unreachable!
%\end{enumerate}
The branching for most of these is trivial: after successful execution the flow of all but \lstinline!br!, \lstinline!ret! and \lstinline!unreachable! will unconditionally attempt to execute the next instruction in memory (i.e. increment the instruction pointer by instruction length).

@@

@@

Dataflow analysis

@@

%% TODO
%* optimistic coloring
\subsection{Coalescing}

@@

\section{Instruction Selection}

@@

\section{Further Optimization}

@@

Benchmarks are performed on Apple Silicon (M1 Pro, i.e. ARMv8.5-A \cite{article}) despite the target instruction set being x86-64. This is possible because of Rosetta 2, which  translates x86-64 to ARMv8 before execution \cite{PolettoMassimiliano1999Lsra}.  Time is measured using the \lstinline!time! function of the bash shell as it is native to the chosen Docker image.  As all execution relevant to the performance of register allocation is done in user mode, only the \lstinline!user! metric is noted.

@@

%TODO:
%
%* should I use shell time, or /usr/bin/time?
%* how do I measure cpu cycles?
\begin{center}
\begin{tabular}{ c c c }
 cell1 & cell2 & cell3 \\ 
 cell4 & cell5 & cell6 \\  
 cell7 & cell8 & cell9    
\end{tabular}
\end{center}

\begin{center}
\begin{tabular}{|c||c|c|c|c|} 
 \hline
 sha256 & -O0 & -O1 & -O2 & -O3 \\ [0.5ex] 
 \hline
 1000000 & 1 & 0.73 & 0.73 & 0.73 \\ [1ex] 
 \hline
\end{tabular}
\end{center}

@@

\section{LLVM-- instruction set}
\label{appendix:llvmis}
\begin{enumerate}
    \item binary operations \lstinline!add!, \lstinline!and!, \lstinline!ashr!, \lstinline!lshr!, \lstinline!mul!, \lstinline!or!, \lstinline!sdiv!, \lstinline!srem!, \lstinline!shl!, \lstinline!sub! and \lstinline!xor!
    \item integer comparison \lstinline!icmp! (conditions being \lstinline!eq!, \lstinline!ne!, \lstinline!sge!, \lstinline!sgt!, \lstinline!sle! and \lstinline!slt!)
    \item memory/address operations \lstinline!alloca!, \lstinline!gep!, \lstinline!load! and \lstinline!store!
    \item  \lstinline!mov! operations \lstinline!gep!, \lstinline!phi!, \lstinline!ptrtoint!, \lstinline!trunc!, and \lstinline!zext!
    \item control flow operations \lstinline!br!, \lstinline!call! and \lstinline!ret!
    \item a  nullary terminator \lstinline!unreachable!
\end{enumerate}
